Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-14 paper.tex
Yves: missing caption

Following the NSA involvement in the design of the \DES{} diffusion
layer~\citep{johnson:cold-war-crypto}, cryptographers have been
advocating for ``nothing-up-my-sleeve'' designs (which does not make
magic tricks impossible, just
harder~\citep{bernstein:something-up-my-sleeves}). For example, the
diffusion layer of the \AES{} standard is based on the usual
multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state modern
ciphers operate on is taken to represent a certain matrix
(oftentimes, a 2D or 3D array of bits): the cipher is not specified
over a flat sequence of individual bits but over a n-dimensional array
of bits.

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidence tends to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 5'' scene to exhibit a
bitsliced witness in all the gory details.

@@

What would it take to turn this art form into an engineering
principle? To answer, we make the following contributions:
%
\begin{itemize}
\item we identify the algebraic structure necessary to model the state
  of modern symmetric cryptographic primitives
  (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
  language of software circuits, rooted in the categorical notion of
  functor to account for data containers~;
\item we give a formal account of bitslicing as a data representation
  change supported by suitably polymorphic definitions
  (Section~\ref{sec:bitslicing}). As expected when it comes to
  switching between data representations, parametricity plays a key
  role to justify the equivalence of the resulting programs~;
\item we give a formal account of fixslicing as a whole-program
  transformation (Section~\ref{sec:fixslicing}). We show how a
  fixsliced implementation can be obtained through equational
  reasoning. To support such reasoning, we crucially rely on the
  algebraic structure set forth in Section~\ref{sec:spec}.
\end{itemize}

@@

For all intents and purposes, these type constructors shall remain
abstract throughout this article: we interact with them solely through
their algebraic interface. The first of which is the categorical
notion of functor:
%
\coqfrom{Algebra}{Functor}

To the disciples of Reynolds and Girard, this signature has come to
mean ``parametric data container'': how would you go about asserting
that the type constructor \coqe!Rows.T! is not doing something
non-trivial with the type it is provided as an argument? In other
words, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
as a \coqe!Rows.T nat!? One could try and argue that Coq does not
allow pattern matching on types but it is rather unsavory to involve
the design of the whole programming language into this argument.

@@

Taking the monoid to be \coqe!list A! (under nil and concatenation,
\ie{} the free monoid construction), we can read off the
directionality we were after.
%
More usefully for our purposes, we can derive generic left and right
rotation operators for any foldable functor \coqe!T! Naperian
over a type \coqe!Ix!:
%
\coqfrom{Rows}{ror_sig}
\coqfrom{Rows}{rol_sig}

@@

Consequently, our hands are tied when it comes to lifting up
individual components:
%
\coqfrom{Skinny/Spec}{round}
%
In particular, we have that \coqe!subcells_! is iterated pointwise
across rows and columns, \coqe!add_round_key_! handles the entire cube
as the support for a Boolean algebra, \coqe!shiftrows_! applicatively
applies its column transformation over individual rows and
\coqe!mixcolumns! handles horizontal faces of the cube as Boolean
algebras.

@@

If we were to adopt \coqe!state! as our actual representation, the
memory usage of the resulting program would be quite poor. It would take
at least 128 bytes to store that many bits (assuming that booleans are
stored in the smallest addressable memory size, a byte). This would be
a terrible waste of memory bandwidth and register usage.

@@

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out of the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure on the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.

>>>>

Following the NSA involvement in the design of the \DES{} diffusion
layer~\citep{johnson:cold-war-crypto}, cryptographers have been
advocating for ``nothing-up-my-sleeve'' designs (which does not make
magic tricks impossible, just
harder~\citep{bernstein:something-up-my-sleeves}). For example, the
diffusion layer of the \AES{} standard is based on the usual
multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state of
modern ciphers operate is taken to represent a certain matrix
(oftentimes, a 2D or 3D array of bits): the cipher is not specified
over a flat sequence of individual bits but over a n-dimensional array
of bits.

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidences tend to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the traditional ``Section 5'' scene to exhibit a
bitsliced witness in all the gory details.

@@

What would it take to turn this art form into an engineering
principle? To answer, we make the following contributions:
%
\begin{itemize}
\item we identify the algebraic structure necessary to model the state
  of modern symmetric cryptography primitives
  (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
  language of software circuits, rooted in the categorical notion of
  functor to account for data containers~;
\item we give a formal account of bitslicing as a data representation
  change supported by suitably polymorphic definitions
  (Section~\ref{sec:bitslicing}). As expected when it comes to
  switching between data representations, parametricity plays a key
  role to justify the equivalence of the resulting programs~;
\item we give a formal account of fixslicing as a whole-program
  transformation (Section~\ref{sec:fixslicing}). We show how a
  fixsliced implementation can be obtained through equational
  reasoning. To support such reasoning, we crucially rely on the
  algebraic structure set forth in Section~\ref{sec:spec}.
\end{itemize}

@@

For all intent and purposes, these type constructors shall remain
abstract throughout this article: we interact with them solely through
their algebraic interface. The first of which is the categorical
notion of functor:
%
\coqfrom{Algebra}{Functor}

To the disciples of Reynolds and Girard, this signature has come to
mean ``parametric data container'': how would you go about asserting
that the type constructor \coqe!Rows.T! is not doing something
non-trivial with the type it is provided as an argument? In other
word, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
as a \coqe!Rows.T nat!? One could try and argue that Coq does not
allow pattern matching on types but it is rather unsavory to involve
the design of the whole programming language into this argument.

@@

Taking the monoid to be \coqe!list A! (under nil and concatenation,
\ie{} the free monoid construction), we can read off the
directionality we were after.
%
More usefully for our purposes, we can derive generic left and right
rotation rotation operators for any foldable functor \coqe!T! Naperian
over a type \coqe!Ix!:
%
\coqfrom{Rows}{ror_sig}
\coqfrom{Rows}{rol_sig}

@@

Consequently, our hands are forced when it comes to lifting up
individual components:
%
\coqfrom{Skinny/Spec}{round}
%
In particular, we have that \coqe!subcells_! is iterated pointwise
across rows and columns, \coqe!add_round_key_! handles the entire cube
as the support for a Boolean algebra, \coqe!shiftrows_! applicatively
applies its column transformation over individual rows and
\coqe!mixcolumns! handles horizontal faces of the cube as Boolean
algebras.

@@

If we were to adopt \coqe!state! as our actual representation, the
memory usage of the resulting program would quite poor. It would take
at least 128 bytes to store that many bits (assuming that booleans are
stored in the smallest addressable memory size, a byte). This would be
a terrible waste of memory bandwidth and register usage.

@@

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure to the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.
